---
title: "ENV S 193DS Final"
format: 
  html:
    toc: true        
    toc-location: left 
    toc-floating: true 
---

```{r setup}
#install packages as needed in console & load in libraries 
library(tidyverse)
library(ggplot2)
library(lme4)
library(DHARMa)
library(MuMIn)
library(ggplot2)
library(dplyr)
library(emmeans) 

# load in data for problems 2 & 3 
sst <- read_csv("../data/SST_SBC.csv") #read one folder up so can find csv in data folder b/c Quarto in code folder 
nest_boxes <- read_csv("../data/occdist.csv") #read in nest occupancy data set, read one folder up

```

# Problem 1. Research writing 

## a. Transparent statistical methods 

In part 1 my co-worker used a correlation test to assess the linear relationship between the distance from the headwater (km) and the annual nitrogen load (kg year⁻¹). In their statement they use the term "correlation" and provide a p-value. In part 2 they used an analysis of variance (ANOVA) to compare the mean nitrogen load across more than one course category (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands). ANOVA is usually sued to test for differences in means cross more than two groups, so we can infer that this is what they likely used.

## b. More information needed 
One test that they could do to provide more context to the results is look into the effect size. Only including a p-value tells us that the difference likely exists but not how large or significant that difference is. By including an effect size would help to assign a quantity for how much of the variation in nitrogen load can be explained by the type of source its coming fro, which is critical for making informed management decisions. Another test that could be done is Tukey's HSD or post-doc pairwise comparison as the results from the ANOVA only tell that some group means differ but doesn't say which ones are different from each other. Doing a post-hoc test would identify whether, for example, nitrogen load from waste water treatment differed significantly than that from fertilizer. 

## c. Suggestions for rewriting
Suggested rewrite for part 1:

As the nitrogen load tends to increase or decrease with distance from the headwater, this suggests a spatial trend in nutrient accumulation or dilution in the waters of the San Joaquin River Delta. This relationship is supported by a correlation analysis (Pearson’s correlation test: r = correlation coefficient, p = 0.03, α = significance level).

Suggested rewrite for part 2:

The average nitrogen load was different among land use and nitrogen source categories, which suggests that certain types of sources contribute more to nitrogen input in the San Joaquin River Delta. This was supported by an analysis of variance (F-statistic = [F-value], degrees of freedom = [df1, df2], p = 0.02, α = [significance level]). Howwver, the inclusion of post-hoc comparisons and effect size measures are needed to identify which sources differ and the magnitude of their contributions.

# Problem 2. Data visualization 

## a. Cleaning and summarizing

```{r}
sst_clean <- sst |>
  # convert date column to date format
  mutate(date = ymd(date)) |>  
  
  # extract year and month
  mutate(
    year = year(date),
    month = month(date, label = TRUE)
  ) |>
  
  # filter to 2018–2023
  filter(year >= 2018, year <= 2023) |>
  
  # group by site, year, and month
  group_by(site, year, month) |>
  
  # summarize mean sea surface temperature
  summarize(
    mean_sst = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )

# view structure and sample of cleaned data
slice_sample(sst_clean, n = 5)
str(sst_clean)

```
## b. Visualize the data 

```{r}
library(ggplot2)
library(dplyr)

# make sure 'month' is ordered correctly
sst_clean$month <- factor(sst_clean$month, levels = month.abb)

# convert year to a factor for correct discrete coloring
sst_clean$year <- factor(sst_clean$year)

# summarize if needed elsewhere
sst_summary <- sst_clean %>%
  group_by(year, month) %>%
  summarise(mean_sst = mean(mean_sst, na.rm = TRUE), .groups = "drop")

# recreate plot
ggplot(sst_clean, aes(x = month, y = mean_sst, group = year, color = year)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c( # lightest to darkest blue
      "2018" = "#ADD8E6",
      "2019" = "#87AFC7",
      "2020" = "#5B7F9E",
      "2021" = "#2F4F4F",
      "2022" = "#1C1C1C",
      "2023" = "#000000"
    ),
    name = "Year"
  ) +
  labs(
    x = "Month",
    y = "Mean monthly sea surface temperature (°C)",
    color = "Year"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),  # remove grid lines
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = c(0.05, 0.85),  # move legend inside top-left
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
  )

```

# Problem 3. Data analysis 

## a. Response variable 

For this data set a 1 represents that a Swift Parrot had occupied a nest during a given season and a 0 means that a nest was unoccupied by Swift Parrots (either occupied by another species or completely unoccupied)

sp		categorical	1 = swift parrot 	0 = other species (or empty)
cs		categorical	1 = common starling 	0 = other species (or empty)
e		categorical	1 = unoccupied (empty) 	0 = occupied
tm		categorical	1 = tree martin		0 = other species


## b. Purpose of study

The main difference between the Swift Parrots and the other bird species in this study is that they are the critically endangered target species in this study, whereas Common Starlings and Tree Martins are competitors. The goal of this study is to assess if nest boxes support the conversation of this species or if they are mostly used by the two competing species. 

## c. Difference in “seasons”

The authors use the term "seasons" to refer to the study years of 2016 and 2019.
These years differ because for the Swift Parrots 2016 was a breeding season and 2019 was not; this reflects the temporal variation in occupancy patterns of the species 

## d. Table of models 

| Model Number | Season Included | Distance to Forest Edge Included | Model Description                    |
|--------------|------------------|----------------------------------|--------------------------------------|
| 1            | No               | No                               | Null model (intercept only)          |
| 2            | Yes              | Yes                              | Saturated model with both predictors |
| 3            | Yes              | No                               | Season only                          |
| 4            | No               | Yes                              | Distance to forest edge only         |

## e. Run the models 
```{r}
#code to run all of models that described would need in d. 

# convert season to factor
nest_boxes$season <- factor(nest_boxes$season)

# model #1 null model (intercept only)
mod1 <- glm(sp ~ 1, data = nest_boxes, family = binomial)

# model #2 saturated model with both predictors
mod2 <- glm(sp ~ season + `edge distance`,  data = nest_boxes, family = binomial)

# model #3 season only 
mod3 <- glm(sp ~ season, data = nest_boxes, family = binomial)

# model #4 Distance to forest edge only
mod4 <- glm(sp ~ `edge distance`, data = nest_boxes, family = binomial)

```

## f. Check the diagnostics
```{r}

#check diagnostic of each model using simulated residuals & display plots 

# model #1
res_null <- simulateResiduals(mod1)
plot(res_null, main = "Diagnostics: Null Model")

# model #2 
res_saturated <- simulateResiduals(mod2)
plot(res_saturated, main = "Diagnostics: Saturated Model")

# model #3 
res_season <- simulateResiduals(mod3)
plot(res_season, main = "Diagnostics: Season-Only Model")

# model #4
res_distance <- simulateResiduals(mod4)
plot(res_distance, main = "Diagnostics: Edge Distance-Only Model")

```
## g. Select the best model 
```{r}
AIC(mod1, mod2, mod3, mod4)
```

The best model as determined by Akaike’s Information Criterion (AIC) was the saturated model (mod2), which included both season and distance to forest edge as predictors with an AIC value of 226.21. This model had the lowest AIC of all four models tested,  which indicates that combining the effects of season and edge distance to the forest best explain the variation in nest occupancy by the Swift Parrots.

## h. Visualize the model predictions 
```{r}

# get predicted probabilities and 95% CI from fitted model 'mod2'
preds <- emmeans(mod2, ~ season * `edge distance`, type = "response")

# convert to data frame to be able to plot 
preds_df <- as.data.frame(preds)


#use prediction data to plot

ggplot() +
  # raw data points
  geom_point(data = nest_boxes,
             aes(x = `edge distance`, y = sp, color = factor(season)), #distance to forest on x-axis & repsonse variable (whether SP in nest or not) on y-axis, do by "year" (called season in dataset)
             alpha = 0.5) +
  
  # predicted probabilities as lines that are grouped by season/year 
  geom_line(data = preds_df,
            aes(x = `edge distance`, y = prob, color = factor(season), group = season),
            size = 1) +
  
  # confidence intervals as ribbons
  geom_ribbon(data = preds_df,
              aes(x = `edge distance`, ymin = asymp.LCL, ymax = asymp.UCL, 
                  fill = factor(season), group = season),
              alpha = 0.2, color = NA) +
  
  # full axis labels and plot title
  labs(
    x = "Distance to Forest Edge (meters)",
    y = "Probability of Nest Box Occupancy",
    color = "Season",
    fill = "Season",
    title = "Swift Parrot Nest Box Occupancy: Model Predictions with 95% Confidence Intervals"
  ) +
  
  # Minimal theme with no gridlines
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  
  # customize colors
  scale_color_manual(values = c("2016" = "turquoise", "2019" = "forestgreen")) +
  scale_fill_manual(values = c("2016" = "turquoise", "2019" = "forestgreen"))

```
## i. Write a caption for your figure. 

Figure 5. Swift Parrot Nest Box Occupancy Probability by Distance to Forest Edge and Season

This figure shows the relationship between the distance to the forest edge and year to the probability of a nest box being occupied by the Swift Parrot. The points represent the observed occupancy data and lines drawn between the points show the model-predicted probabilities taken from Model #2; shaded ribbons are supposed to represent the corresponding 95% confidence intervals. This plot illustrates that there is a decline in occupancy probability as distance from the forest edge increases as we see seasonal differences represented by there being separate prediction lines for each year (Data source: Adapted from Stojanovic et al. (2021), "Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird." Biological Conservation, 258, 109136) 

## j. Calculate model predictions 

```{r}

# define specific values of 'edge distance' that want predictions for (just 0 to 900 m from forest edge)
new_edge_distances <- c(0, 900)

# get model predictions at these points, for each season
pred_points <- emmeans(mod2, 
                      ~ season,                 # predict by season/year
                      at = list(`edge distance` = new_edge_distances),  # at 0 m and at 900 m
                      type = "response")        # return predictions on response scale (probabilities)

# convert to data frame so can view it easier 
pred_points_df <- as.data.frame(pred_points)

# display results
print(pred_points_df)

```
## k. Interpret your results 

The figure made in part h (Figure 5) shows the model predictions of the probability of nest box occupancy by the Swift Parrot as a function of distance from the forest edge, with separate lines for each season (2016 and 2019). The predicted probabilities calculated at specific distances (0 to 900 m) made in part j reveal that occupancy is higher at the forest edge and as distance from the forest edge increased to 900 m, the probability of occupancy decreased rapidly; this indicates a significant negative relationship between distance and the probability of occupancy. This trend likely reflects the species' preference or dependence on forest edge habitats for food availability and protection from predation and differences between the two seasons analyzed may reflect annual variations in species population or environmental conditions that affect the suitability of the habitat. This trend is also in opposition to a comperting species examined in the study, the Tree Martin, who's occupancy of nest boxes increases with their distance from the forest edge in both 2016 and 2019. 

# Problem 4. Affective and exploratory visualizations

## a. Comparing visualizations

The exploratory visualization I made for Homework 2 is much more traditional in the sense that it is a box-plot that shows the correlation between productivity level and the type of music listened to for a handful of study session data entries. The affective visualization made for Homework 3 tells a more complete story about how each type of music affected my studying through the use of color, notes about my emotions and mental state, and differing sizes of brain waves 

Between the visualizations I would say there are similarities in that greater productivity is associated more closely with listening to classical or ambient music. Both also show that I am less focused and productive when listening to elctropop and folk music 

Homework 2 contains less data entries to work off of than Homework 3 but overall the trend stayed the same; greater productivity and focus is associated with classical and ambient music and more distraction is associated with electropop and folk music. I would say that it is hard to make definitive conclusions, as there are more instances where I am listening to either classical or ambient music while studying than when I choose to listen to electropop or folk music, meaning that the sample size for these music types is greater and more predictable. 

I got some feedback about how to orient the visualization as I was having trouble figuring out how to do a horizontal color gradient and a peer recommended I try orienting it diagonally across my paper which worked better. Also, this wasn't a recommendation from a classmate or instrcutor, but I decided to pivot my visualization a bit from my initail sketch to use brain waves as the representation of the focus level, which I feel worked better to convey how I was feeling in each study session based on the music type 

## b. Sharing your affective visualization 

Attendance to workshop 10 during Week 10 :)